```{r loadlib,echo=FALSE,results='hide'}
library(lattice)
```


# Reproducible Research: Peer Assessment 1


## Introduction
This file is the R Mark down file for Reproducible Research peer assessment 1. Probelm details regarding this assessment is available in [Read Me file]("https://github.com/vinaykk/RepData_PeerAssessment1/blob/master/README.md") 

Following are the sections related to problems asked in this assessment. 

## Loading and preprocessing the data

Load the activity data
```{r load activity data}
DFActivity <- read.csv("activity.csv",colClasses=c("numeric","Date","numeric"))
```
Process/transform data
```{r transform data for mean calculation}
DFTotalStepsPerDay <- aggregate(steps ~ date, data = DFActivity , FUN="sum")
DFAvgStepsPerInt <- aggregate(steps ~ interval, data = DFActivity , FUN="mean")
```


## What is mean total number of steps taken per day?

Generate histogram of the total number of steps taken each day
```{r plot histogram for mean total number of steps taken per day}
hist(DFTotalStepsPerDay$steps,col="Red",freq=TRUE,main="Total number of steps taken each day",xlab="Total Steps")
```

Calculate and report the mean and median total number of steps taken per day
```{r calc mean & median}
stepsMean <- mean(DFTotalStepsPerDay$steps)

stepsMedian <- median(DFTotalStepsPerDay$steps)
```

The Mean total number of steps taken per day is `r format(stepsMean)`    
The Median total number of steps taken per day is `r format(stepsMedian)`

## What is the average daily activity pattern?

Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

```{r plot time series for average daily activity pattern}
plot(DFAvgStepsPerInt$interval,DFAvgStepsPerInt$steps,type="l" )
````

Calculate 5- Minute interval that has max avergate across all day. 
```{r calc 5-Minute interval}
max5MinInt <- DFAvgStepsPerInt[ (DFAvgStepsPerInt$steps  == max(DFAvgStepsPerInt$steps)),]$interval
```

`r format(max5MinInt-5)` - `r format(max5MinInt)` is the 5-minute interval, on average across all the days in the dataset, containg the maximum number of steps.

## Imputing missing values
Note that there are a number of days/intervals where there are missing values (coded as NA). The presence of missing days may introduce bias into some calculations or summaries of the data.

Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r missing values}
ncnt <- summary(is.na(DFActivity$steps))[3][1]
```
Total number of missing values in the dataset is `r ncnt`

Following is the strategy used to replace missing values. 
1. Make copy of data set
2. Calculate mean steps for each  5-minute interval 
3. Replace NA's with calculated mean

```{r Apply strategy to replace missing values}
DFActivityImput <- DFActivity
DFActivityImput$step[(is.na(DFActivityImput$steps))] <- round(ave(DFActivityImput$steps, DFActivityImput$interval, FUN = function(x) mean(x, na.rm=T) )[(is.na(DFActivityImput$steps))])
```

Make a histogram of the total number of steps taken each day
```{r plot histogram mean total number of steps taken per day using imputed data}
DFSumStepsPerDayImput <- aggregate(steps ~ date, data = DFActivityImput , FUN="sum")

hist(DFSumStepsPerDayImput$steps,col="Red",freq=TRUE,main="Total number of steps taken each day",xlab="Total Steps")
```

Calculate and report the mean and median total number of steps taken per day
```{r calc mean & median using imputed data}
stepsMean <- mean(DFSumStepsPerDayImput$steps)

stepsMedian <- median(DFSumStepsPerDayImput$steps)
```

The Mean total number of steps taken per day is `r format(stepsMean)`
The Median total number of steps taken per day is `r format(stepsMedian)`

Mean and Median total number of steps taken does not differ after applying strategy to replace missing steps count with mean steps of 5-minute interval.

Estimates of the total daily number of steps has no impact after applying stategy to replace missing steps count with mean steps of 5-minute interval. 


## Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

```{r create weekday/weekend Factor variable}

DFActivity$DayType <- as.factor( ifelse( (weekdays(DFActivity$date) == "Sunday" | weekdays(DFActivity$date) == "Saturday"),"weekday","weekend" ))

DFAvgStepsPerInt <- aggregate(steps ~ interval + DayType, data = DFActivity , FUN="mean")
```

```{r plot pannels for differences in activity patterns between weekdays and weekends}
xyplot(steps ~ interval | DayType, data = DFAvgStepsPerInt,type="l", layout = c(1,2), ylab="Number of Steps", xlab="Interval" )

```



